# Volatility modeling {#session06}

## GARCH(1,1) estimation

In this session, we implement the maximum likelihood estimation of the GARCH(1,1) model. We consider a Gaussian GARCH(1,1)

$$
\begin{aligned}
r_t &= \sigma^2_t z_t \\[1ex]
\sigma^2_t &= \omega + \alpha r_{t-1}^2 + \beta \sigma^2_{t-1}
\end{aligned}
$$

where $z_t \sim \mathcal{N}(0, 1)$. The log-likelihood is (Exercise 3):

$$
\ell(\theta; r_1, \dots, r_T) = -\frac{T-1}{2}\log(2\pi) - \frac{1}{2}\sum_{t=2}^T\log\hat\sigma_t -\frac{1}{2} \sum_{t=2}^T \frac{r_t^2}{\hat\sigma^2_t}
$$

The values of the latent volatility $\{\sigma_2^2, \sigma_3^2, \dots, \sigma_T^2\}$ can be estimated using the following filtering procedure for each given value $\theta = (\omega, \alpha, \beta)$:

 * set $\hat\sigma^2_t = \hat{\text{Var}}(r_t)$
 * for $t = 2, \dots, T$ set $\hat\sigma^2_t = \omega + \alpha r_{t-1}^2 + \beta \hat\sigma^2_{t-1}$ 

We simulate the data according to the GARCH(1,1) process and compare the MLE with the true parameters.


```{r}
## Simulate data
t_max <- 1000
w <- 0.01
alpha <- 0.05
beta <- 0.949

z <- rnorm(t_max)
sigma <- c(1, rep(NA, t_max))
r     <- c(sigma[1]*z[1], rep(NA, t_max))
for (t in 2:(t_max+1)) {
  sigma[t] <- w + alpha*r[t-1] + beta*sigma[t-1]
  r[t]     <- sigma[t]*z[t]
}
r <- r[100:1000]
```

```{r}
plot.ts(r)
plot.ts(r^2)
```




Estimation


```{r}
## filter 
garch11_filter <- function(data, params){
  # initializations
  omega <- params[1]; 
  alpha <- params[2]; 
  beta <- params[3];
  T <- length(data);
  sig2 <- numeric(T); 
  z <- numeric(T); 
  loglik <- 0
  sig2[1] <- var(data); z[1] <- data[1]/sqrt(sig2[1])
  # for loop calculating one-step-ahead
  for (t in 2:T){
    sig <- omega + alpha * data[t-1]^2 + beta * sig2[t-1]
    sig2[t] <- sig; z[t] <- data[t]/sqrt(sig)
  }
  # loglik
  loglik <- -T/2*log(2*pi) - sum(log(sig2))/2 - sum(data^2/sig2)/2
  return(list(sig2 = sig2, z = z, loglik = loglik))
}
```


```{r}
## likelihood
garch11_objective <- function(data, params) {
  
  #initializations
  omega <- params[1]; alpha <- params[2]; beta <- params[3]
  T <- length(data)
  sig2 <- garch11_filter(data, params)$sig2
  
  # if-else statement calculating negative loglik
  if (all(is.finite(params)) & omega > 0 & alpha > 0 & beta > 0 & alpha + beta < 1) {
    neg_loglik <- -(-T/2*log(2*pi) - sum(log(sig2))/2 - sum(data^2/sig2)/2)
  } else {
    neg_loglik <- Inf
  }
  return(neg_loglik)
}
```


```{r}
## MLE
garch11_mle <- function(data, params) {
  # nlminb function optimizing parameters 
  fit <- nlminb(start = params, objective = garch11_objective, 
                data = data, lower = 0.0001, upper = 1)
  return(param_mle = fit$par)
}
```


```{r}
mle <- garch11_mle(r, c(0, 0.1, 0.1))
```

```{r}
cbind(mle, c(w, alpha, beta))
```



## Exercises

**Exercise 1 (ARCH($\infty$) representation of GARCH(1,1))**

Show that the GARCH(1,1)

$$
\begin{aligned}
r_t &= \sigma^2_t z_t \\[1ex]
\sigma^2_t &= \omega + \alpha r_{t-1}^2 + \beta \sigma^2_{t-1}
\end{aligned}
$$
with $\mathbb{E}[z_t] = 0$, $\mathbb{E}[z^2_t] = 1$ can be rewritten as an ARCH($\infty$) with

$$
\sigma^2_t = \frac{\omega}{1-\beta} + \alpha \sum_{j=0}^\infty \beta^j \ r^2_{t-j-1}
$$

under the assumption that $|\beta| < 1$.


<br>
<br>

**Exercise 2 (Autocovariance of the GARCH(1,1))**

Consider the GARCH(1,1)

$$
\begin{aligned}
r_t &= \sigma^2_t z_t \\[1ex]
\sigma^2_t &= \omega + \alpha r_{t-1}^2 + \beta \sigma^2_{t-1}
\end{aligned}
$$

with $\mathbb{E}[z_t] = 0$, $\mathbb{E}[z^2_t] = 1$.


Show that:

$$
\gamma(k) = \text{cov}(r_t, r_{t-k}) = 
\begin{cases}
\frac{\omega}{1-\alpha-\beta}  \qquad  & k = 0 \\
0 & k \neq 0
\end{cases}
$$


*[Difficult question]* Also show that

$$
\gamma_2(k) = \text{cov}(r^2_t, r^2_{t-k}) = 
\begin{cases}
\text{Var}(r^2_t - \sigma^2_t)\left[1+\frac{\alpha+2\beta}{1-(\alpha+\beta)^2}\right] \qquad  & k = 0 \\[1em]
\text{Var}(r^2_t - \sigma^2_t)\left[\alpha+2\beta+\frac{(\alpha+2\beta)^2 (\alpha+\beta)}{1-(\alpha+\beta)^2}\right] \qquad  & k = 1 \\[1em]
(\alpha+\beta)^{k-1}\gamma(1) & k \geq 2
\end{cases}
$$
where $\text{Var}(r^2_t - \sigma^2_t) = (\mu_4 - 1)\left( \sigma_4 - \left(\frac{\omega}{1-\alpha-\beta}\right)^2 \right)$, $\sigma^4 = \mathbb{E}[\sigma^4_t]$ (you do not need to find the explicit expression for this in terms of the parameters) and $\mu_4 = \mathbb{E}[z_t^4]$.

**Hint:** Show that the GARCH(1,1) admits an ARMA(1,1) representation $\sigma_t^2 = \omega + (\alpha+\beta)r^2_{t-1} + \beta(r^2_{t-1}-\sigma^2_{t-1}) + (r^2_{t}-\sigma^2_{t})$ and apply the formula for the autocovariance of the ARMA(1,1) (see Exercise 3 in Session 4).



<br>
<br>

**Exercise 3 (Log-likelihood of the GARCH(1,1))**
Consider a Gaussian GARCH(1,1)

$$
\begin{aligned}
r_t &= \sigma^2_t z_t \\[1ex]
\sigma^2_t &= \omega + \alpha r_{t-1}^2 + \beta \sigma^2_{t-1}
\end{aligned}
$$

where $z_t \sim \mathcal{N}(0, 1)$. Show that the conditional log-likelihood of the model is

$$
\ell(\theta; r_1, \dots, r_T) = -\frac{T-1}{2}\log(2\pi) - \frac{1}{2}\sum_{t=2}^T\log\hat\sigma_t -\frac{1}{2} \sum_{t=2}^T \frac{r_t^2}{\hat\sigma^2_t}
$$

where $\hat\sigma_t^2$ is the predicted value of the latent volatility computed using the filtering procedure described in the session.


<!-- # Sharing your book -->

<!-- ## Publishing -->

<!-- HTML books can be published online, see: https://bookdown.org/yihui/bookdown/publishing.html -->

<!-- ## 404 pages -->

<!-- By default, users will be directed to a 404 page if they try to access a webpage that cannot be found. If you'd like to customize your 404 page instead of using the default, you may add either a `_404.Rmd` or `_404.md` file to your project root and use code and/or Markdown syntax. -->

<!-- ## Metadata for sharing -->

<!-- Bookdown HTML books will provide HTML metadata for social sharing on platforms like Twitter, Facebook, and LinkedIn, using information you provide in the `index.Rmd` YAML. To setup, set the `url` for your book and the path to your `cover-image` file. Your book's `title` and `description` are also used. -->



<!-- This `gitbook` uses the same social sharing data across all chapters in your book- all links shared will look the same. -->

<!-- Specify your book's source repository on GitHub using the `edit` key under the configuration options in the `_output.yml` file, which allows users to suggest an edit by linking to a chapter's source file.  -->

<!-- Read more about the features of this output format here: -->

<!-- https://pkgs.rstudio.com/bookdown/reference/gitbook.html -->

<!-- Or use: -->

<!-- ```{r eval=FALSE} -->
<!-- ?bookdown::gitbook -->
<!-- ``` -->


