# Volatility modeling in practice {#session07}


## The ARMA(1,1)-TGARCH(1,1) model

In this session we include the conditionally-heteroskedastic models into the ARMA framework. In particular, we consider an ARMA(1,1) model with conditionally-heteroskedastic innovations and asymmetric effects. For simplicity, we just consider first-order difference equations, but the model can be generalized to include more lags. The model is an ARMA(1,1)-TGARCH(1,1) of the following form:


$$
\begin{aligned}
y_t &= c + \phi y_{t-1} + \theta \varepsilon_{t-1} + \varepsilon_t \\
\varepsilon_t &= \sigma_t z_t \\
\sigma^2_t &= \omega + \alpha y_{t-1}^2 + \beta \sigma^2_{t-1} + \gamma y_{t-1}^2 \mathbb{1}_{y_{t-1}<0}
\end{aligned}
$$

Throughout this session, we consider $z_t \sim \mathcal{N}(0,1)$ and we estimate the model via maximum-likelihood. Let $I_t = \{y_1, \dots, y_t, \varepsilon_1, \dots, \varepsilon_t, \sigma_1, \dots, \sigma_t\}$. The log-likelihood of the model is:

$$
\ell(\theta; I_T) = -\frac{T-1}{2} \log(2\pi) - \frac{1}{2}\sum_{t=2}^T \log \sigma_t(I_{t-1}; \theta) - \frac{1}{2}\sum_{t=2}^T \frac{( \varepsilon_t(I_{t-1}; \theta))^2}{\sigma^2_t(I_{t-1}; \theta)},
$$

where expressions for $\sigma_t(I_{t-1}; \theta), \ \varepsilon_t(I_{t-1}; \theta)$ can be found using a standard filtering procedure (Exercise 3).

```{r}
## Simulate data from ARMA(1,1)-TGARCH(1,1) process
# set parameters
t_max <- 1000
c     <- 0
phi   <- 0.85
theta <- -0.1
omega <- 0.01
alpha <- 0.1
beta  <- 0.85
gamma <- 0.05

# simulate
set.seed(123)
z <- rnorm(t_max)
epsilon <- c(0, rep(NA, t_max))
sigma2 <- c(1, rep(NA, t_max))
y     <- c(sigma2[1]*z[1], rep(NA, t_max))
for (t in 2:(t_max+1)) {
  set.seed(t*212) # for reproducibility
  # sigma2[t]  <- omega + alpha*y[t-1]^2 + beta*sigma2[t-1] + gamma*y[t-1]^2 + gamma*y[t-1]^2*(y[t-1]<0)
  sigma2[t]  <- omega + alpha*y[t-1]^2 + beta*sigma2[t-1] + gamma*y[t-1]^2*(y[t-1]<0)
  epsilon[t] <- sigma2[t]*z[t]
  y[t]       <- c + phi*y[t-1] + theta*epsilon[t-1] + epsilon[t]
}
y <- y[50:t_max]
```

```{r}
# plot simulated series
plot.ts(y)

# plot simulated series squared
plot.ts(y^2)
```



```{r}
## filter 
arma11tgarch11_filter <- function(y, params){
  # initializations
  c      <- params[1]
  phi    <- params[2]
  theta  <- params[3]
  omega  <- params[4]
  alpha  <- params[5]
  beta   <- params[6]
  gamma  <- params[7]
  t_max  <- length(y)
  
  eps  <- c(0, rep(NA, t_max-1))
  sig2 <- c(1, rep(NA, t_max-1))
  z    <- rep(NA, t_max)
  loglik <- 0
  sig2[1] <- var(y)
  z[1]    <- y[1]/sqrt(sig2[1])
  # for loop calculating one-step-ahead
  for (t in 2:t_max){
    eps[t]  <- y[t] - c - phi*y[t-1] - theta*eps[t-1]
    sig2[t] <- omega + alpha*y[t-1]^2 + beta*sig2[t-1] + gamma*y[t-1]^2*(y[t-1]<0)
    z[t]    <- eps[t]/sqrt(sig2[t])
  }
  # loglik
  loglik <-  - sum(log(sig2[2:t_max])) - sum(eps[2:t_max]^2/sig2[2:t_max])  # *0.5 + constant terms 
  
  # output
  return(list(sig2   = sig2, 
              eps    = eps,
              z      = z, 
              loglik = loglik))
}
```



```{r}
## likelihood
arma11tgarch11_objective <- function(y, params) {
  #initializations
  c      <- params[1]
  phi    <- params[2]
  theta  <- params[3]
  omega  <- params[4]
  alpha  <- params[5]
  beta   <- params[6]
  gamma  <- params[7]
  t_max  <- length(y)
  res_filter <- arma11tgarch11_filter(y, params)
  sig2 <- res_filter$sig2
  eps <- res_filter$eps
  
  # if-else statement calculating negative loglik
  # if (all(is.finite(params)) & omega>=0 & alpha>0 & beta>0 & gamma>0 & (alpha+beta+gamma/2)<1) {
  if (all(is.finite(params)) & (alpha+beta+gamma/2)<1) {
    neg_loglik <- -res_filter$loglik
  } else {
    neg_loglik <- Inf
  }
  return(neg_loglik)
}
```

```{r include = FALSE}
# nlm(arma11tgarch11_objective, p = params, y = y, )
```


```{r}
## MLE
arma11tgarch11_mle <- function(y, params) {
  # nlminb function optimizing parameters 
  fit <- nlminb(start = params, objective = arma11tgarch11_objective, 
                y = y, 
                lower = c(-1, -0.999, -0.999, -1, 0.01, 0.01, 0.001), 
                upper = c( 1,  0.999,  0.999,  1, 0.9, 0.9, 0.9))

  ## output
  names(fit$par) <- c("c", "phi", "theta", "omega", "alpha", "beta", "gamma")
  fit$par
}
```

```{r message=FALSE, warning=FALSE}
startparams <- c(0, 0.9, 0, 0.1, 0.1, 0.8, 0.01)
mle <- arma11tgarch11_mle(y, startparams)
```

```{r}
tab <- cbind(MLE = mle, 
             true = c(c, phi, theta, omega, alpha, beta, gamma))
row.names(tab) <- c("c", "phi", "theta", "omega", "alpha", "beta", "gamma")
round(tab, 3)
```

## Modeling the S&P 500 volatility

We now show an application of the ARMA(1,1)-TGARCH(1,1) model to the S&P 500 historical data. 

```{r}
## Load data
sp500 <- read.csv("../data/sp500.csv")
sp500 <- ts(sp500$adjusted_close[nrow(sp500):1], start = c(1999, 11, 1), freq = 365)
d <- diff(log(sp500))*252
plot.ts(d, main = "S&P 500 annualized growth rates")
```

```{r eval = FALSE}
## Compute MLE
mle_sp500 <- arma11tgarch11_mle(d, startparams)
```


```{r eval = TRUE, include = FALSE}
# saveRDS(mle_sp500, file = "../savedobjects/session07_mleSP500.RDS")
mle_sp500 <- readRDS(file = "../savedobjects/session07_mleSP500.RDS")
```

```{r}
mle_sp500
```

```{r}
## plot fitted values
filter_sp500 <- arma11tgarch11_filter(d, mle_sp500)
eps  <- filter_sp500$eps
sig2 <- filter_sp500$sig2
z    <- filter_sp500$z

c      <- mle_sp500[1]
phi    <- mle_sp500[2]
theta  <- mle_sp500[3]
omega  <- mle_sp500[4]
alpha  <- mle_sp500[5]
beta   <- mle_sp500[6]
gamma  <- mle_sp500[7]
t_max  <- length(d)

fitted_sp500 <- c + phi*d + theta*eps
upper_sp500     <- fitted_sp500 + 1.96*sqrt(sig2)
lower_sp500     <- fitted_sp500 - 1.96*sqrt(sig2)
```

```{r}
plot.ts(d, main = "S&P 500 annualized growth rates")
lines(fitted_sp500, col = "steelblue")
lines(lower_sp500, col = "tomato")
lines(upper_sp500, col = "tomato")
text(x = 2009, y = 30,
     labels = paste0("Observations outside CI: ", 
                   round(mean((d > upper_sp500) | (d < lower_sp500)),4)*100, "%"))
```



## Exercises

**Exercise 1 (h-step ahead volatility forecast of the GARCH(1,1))**

Consider the GARCH(1,1) model

$$
\begin{aligned}
r_t &= \sigma^2_t z_t \\[1ex]
\sigma^2_t &= \omega + \alpha r_{t-1}^2 + \beta \sigma^2_{t-1}
\end{aligned}
$$

Show that 

$$
\mathbb{E}[\sigma^2_{t+h} | r_t, \sigma_t] = 
\begin{cases}
\omega + \alpha r_{t}^2 + \beta \sigma^2_t & \qquad \text{if} \quad h=1 \\
\omega \frac{1-(\alpha+\beta)^{h-1}}{1-\alpha-\beta} + (\alpha+\beta)^{h-1} \
\mathbb{E}[\sigma^2_{t+1} | r_t, \sigma_t]  & \qquad \text{if} \quad h>1
\end{cases}
$$

<br>
<br>

**Exercise 2 (Log-likelihood of the ARMA(p,q)-GARCH(r,s) model)**

Consider the following specification of an ARMA model with conditionally-heteroskedastic innovations:

$$
\begin{aligned}
y_t &= c + \phi_1 y_{t-1} + \dots + \phi_p y_{t-p} + \theta_1 \varepsilon_{t-1} + \dots + \theta_q \varepsilon_{t-q} + \varepsilon_t \\
\varepsilon_t &= \sigma_t z_t \\
\sigma^2_t &= \omega + \alpha_1 y_{t-1}^2 + \dots + \alpha_r y_{t-r}^2 + \beta_1 \sigma^2_{t-1} + \dots + \beta_s \sigma^2_{t-s}
\end{aligned}
$$

Derive the log-likelihood of the model under the assumption that $z_t \sim \mathcal{N}(0, 1)$. Provide also the filtering procedure used to estimate the latent variables $\varepsilon_t$ and $\sigma_t$.

<br>
<br>

**Exercise 3 (Log-likelihood of the ARMA(1,1)-TGARCH(1,1) model)**


Consider the following specification of an ARMA model with asymmetric conditionally-heteroskedastic innovations:

$$
\begin{aligned}
y_t &= c + \phi y_{t-1} + \theta \varepsilon_{t-1} + \varepsilon_t \\
\varepsilon_t &= \sigma_t z_t \\
\sigma^2_t &= \omega + \alpha y_{t-1}^2 + \beta \sigma^2_{t-1} + \gamma y_{t-1}^2 \mathbb{1}_{y_{t-1}<0}
\end{aligned}
$$

Derive the log-likelihood of the model under the assumption that $z_t \sim \mathcal{N}(0, 1)$. Provide also the filtering procedure used to estimate the latent variables $\varepsilon_t$ and $\sigma_t$.

<br>
<br>












<!-- `r if (knitr::is_html_output()) ' -->
<!-- # References {-} -->
<!-- '` -->
